Online Verification of Commutativity

### 2020-10-16 acmPaper/paper.tex
tex cleanup; everythong passing except font type 3

\newcount\draft\draft=0 % set to 0 for submission or publication
% \documentclass[sigplan,review,anonymous]{acmart}
% \documentclass[sigplan,nonacm=true]{acmart}
\documentclass[sigplan]{acmart}
%

\usepackage{tikz}
\usetikzlibrary{snakes}
\usepackage[norelsize]{algorithm2e}
\usepackage{xxx}
\usepackage{fncylab}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{microtype}
\usepackage{float}
% \usepackage{endfloat}
% \usepackage{afterpage}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

\begin{document}
\settopmatter{printacmref=true}
\title{Online Verification of Commutativity}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Aditi Kabra}
\affiliation{%
  \institution{Carnegie Mellon University}
  \city{Pittsburgh}
  \state{PA}
  \country{USA}
}
\email{akabra@cs.cmu.edu}

@@

\begin{abstract}
Systems of transformations arise in many programming systems, such as in graphs of implicit type conversion functions.
It is important to ensure that these diagrams commute: that composing any path of transformations from the same source to the same destination yields the same result.
However, a straightforward approach to verifying commutativity must contend with cycles, and even so it runs in exponential time.
Previous work has shown how to verify commutativity in the special case of acyclic diagrams in $O(|V|^4|E|^2)$ time,
% I think this is clear enough (at least for the abstract). --AS
% where V is the set of vertices in the diagram, and E, the set of edges.
but this is a \emph{batch} algorithm: the entire diagram must be known ahead of time.
We present an \emph{online} algorithm that efficiently verifies that a commutative diagram remains commutative when adding a new edge.
The new incremental algorithm runs in $O(|V|^2(|E| + |V|))$ time.
For the case when checking the equality of paths is expensive, we also present an optimization that runs in $O(|V|^4)$ time but reduces to the minimum possible number of equality checks.
We implement the algorithms and compare them to batch baselines, and we demonstrate their practical application in the compiler of a domain-specific language for geometry types.
To study the algorithms' scalability to large diagrams, we apply them to discover discrepancies in currency conversion graphs.
\end{abstract}

@@

A path is a sequence of edges. The edge-to-function mapping $f$ can be naturally extended to paths: if path $p=e_1 ; e_2 ; \cdots ; e_n$ then $f(p)=f(e_1) ; f(e_2) ; \cdots ; f(e_n)$.
%
We write $\partial(p)^{+}$ for $p$'s start node, $\partial(p)^{-}$ for its end node, and $\partial(p)$ to denote the pair
$(\partial(p)^{+}, \partial(p)^{-})$.

@@

parallelPairs $\gets$ \{\}\;
\For{src in existingGraph.Nodes}{
    \For{snk in existingGraph}{
        \try{}{
            \tcc{Use any standard path finding algorithm such as BFS to find a path in the existing graph from the specified source to sink.}
            Path pathWithNewEdge $\gets$ FindPath(
                % graph: existingGraph, 
                sourceNode: src,
                sinkNode: newEdge.Source) +\;
            newEdge +\;
            FindPath(
                % graph: exisingGraph, 
                sourceNode: newEdge.Sink, 
                sinkNode: snk)\;
            \If{src == snk}{
                \tcc{Assign the nullary path from src to snk.}
                pathInOldGraph $\gets$ src\;
            }
            \Else(){
                Path pathInOldGraph $\gets$ FindPath(
                    % graph: existingGraph, 
                    sourceNode: src, 
                    sinkNode: snk)\;
            }
            parallelPairs.add((pathInOldGraph, pathWithNewEdge))\;
        }
        \catch{PathFindingFailedException}{
           \tcc{No comparable pairs from node src to node snk that need to be checked}
           continue\;
        }
    }
}
\KwRet{parallelPairs}\;
\caption{Online polynomial time algorithm to find parallel pair set.}
\label{algo_online_polynomial}
\end{algorithm}

@@

We conclude that verifying a comparable pair of paths with end points ($P_1$, $P_2$) implies the verification of all path pairs ($Q_1$, $Q_2$) such that $Q_1$ is a successor of $P_1$ and $P_2$ is a successor of $Q_2$. A successor $S$ to node $N$ is any node such that there exists a path from $N$ to $S$. Nodes are also their own successors and predecessors.
The rule effectively places an ordering on the informativeness of path pairs based on their terminal nodes.
% If it is possible to draw a path from node $P_1$ to $P_2$ that passes through $Q_1$ and $Q_2$ then verifying a parallel pair from $Q_1$ to $Q_2$ implies that parallel pairs from $P_1$ to $P_2$ commute as well.

@@

\begin{table}
    \caption{Computation time for 9-node graph of density 0.4, averaged over ten runs.}
\begin{tabular}{cc}
    \toprule
    Algorithm & Average seconds of computation \\
    \midrule
    Na\"{i}ve baseline & 0.77 \\
    Two Flip tolerant & 0.075 \\
    Batch algorithm & 7.55 \\
    ~\ref{algo_online_polynomial} & 0.0038 \\
    ~\ref{algo_online_minimal} & 0.00086 \\
    \bottomrule
\end{tabular}
\label{tab:times} %TODO debug and fix label not working.
\end{table}

\begin{figure*}
    % TODO format nicely.
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_OptimalSet.pdf}
      \caption{~\ref{algo_online_minimal}.}
      \label{fig:sfigOptimalTvsS}
    \end{subfigure}
    %
    \hspace{-20pt}
    %
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_Polynomial.pdf}
      \caption{~\ref{algo_online_polynomial}.}
      \label{fig:sfigPolynomialTvsS}
    \end{subfigure}
    %
    \hspace{-20pt}
    %
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_NaiveChecker.pdf}
      \caption{Naive baseline.}
      \label{fig:sfigNaiveTvsS}
    \end{subfigure}

% \end{figure}
% \begin{figure}\ContinuedFloat

@@

@@

We define density to be the ratio of the number of edges in the graph to the total possible number of edges (which is $|V|^2$, where $|V|$ is the number of nodes).
Run time relates to the density of edges in the input graph.
The degree of the effect differs with the algorithms, as Figure~\ref{fig:timeVsSize} shows.
Generally, denser graphs entail longer computation time.
For the batch algorithm we use lower densities since the input graph must be acyclic. This puts an upper bound on density that approaches 0.5 in large graphs.
% Our proposed algorithms, ~\ref{algo_minimal_spanning_set} and ~\ref{algo_online_polynomial} are less affected because they aggressively remove redundancies.

@@

\begin{figure}
    % TODO format nicely.
    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{variance_10_OptimalSet.pdf}
      \caption{~\ref{algo_online_minimal}.}
      \label{fig:sfigOptimalSpread}
    \end{subfigure}

    % \begin{subfigure}{\linewidth}
    %   \includegraphics[width=\linewidth]{variance_10_TwoFlip.pdf}
    %   \caption{Two flip tolerant baseline}
    %   \label{fig:sfigTwoFlipSpread}
    % \end{subfigure}

@@

Section ~\ref{sec:batchBaseline} describes Murota's solution to efficiently finding the minimal set of path pairs that need to be compared to check if a given acyclic graph commutes ~\cite{commutative}.
We did not find any other work that solves the question of verifying that diagrams commute.

However, the question of commuting does come up in programming languages with implicit type conversion.
Gator ~\cite{gator}, as described in Section ~\ref{sec:gator}, supports automatic type conversion between geometry types.
The language implements some restrictions to eliminate obvious cases of non-commuting graphs, but does not verify that defined graphs commute, allowing scope for non-commuting graph definitions.
Frink ~\cite{frink} is a language that supports automatic conversion between units and infinite precision floating point numbers.
It does not appear to support the implicit definition of conversion between units \xxx[AK]{Need to spend some time confirming this} but if extended to do so, would need to contend with the problem of commuting graphs.
The same is true for F\# which has support for units of measure ~\cite{unitsOfMeasure} and Ada's GNAT compiler~\cite{gnat}.

% Units of measure work resource: https://gmpreussner.com/research/dimensional-analysis-in-programming-languages
\section{Conclusion}

>>>>

\newcount\draft\draft=0 % set to 0 for submission or publication
\documentclass[sigplan, scree, balance]{acmart}
%

\usepackage{tikz}
\usetikzlibrary{snakes}
\usepackage[norelsize]{algorithm2e}
\usepackage{xxx}
\usepackage{fncylab}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{microtype}
\usepackage{float}
\usepackage{balance}
%
\usepackage{graphicx}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

\begin{document}
\settopmatter{printacmref=true}
\title{Online Verification of Commutativity}

\author{Aditi Kabra}
\affiliation{%
  \institution{Carnegie Mellon University}
  \city{Pittsburgh}
  \state{PA}
  \country{USA}
}
\email{akabra@cs.cmu.edu}

@@

\begin{abstract}
Systems of transformations arise in many programming systems, such as in graphs of implicit type conversion functions.
It is important to ensure that these diagrams commute: that composing any path of transformations from the same source to the same destination yields the same result.
However, a straightforward approach to verifying commutativity must contend with cycles, and even so it runs in exponential time.
Previous work has shown how to verify commutativity in the special case of acyclic diagrams in $O(|V|^4|E|^2)$ time,
but this is a \emph{batch} algorithm: the entire diagram must be known ahead of time.
We present an \emph{online} algorithm that efficiently verifies that a commutative diagram remains commutative when adding a new edge.
The new incremental algorithm runs in $O(|V|^2(|E| + |V|))$ time.
For the case when checking the equality of paths is expensive, we also present an optimization that runs in $O(|V|^4)$ time but reduces to the minimum possible number of equality checks.
We implement the algorithms and compare them to batch baselines, and we demonstrate their practical application in the compiler of a domain-specific language for geometry types.
To study the algorithms' scalability to large diagrams, we apply them to discover discrepancies in currency conversion graphs.
\end{abstract}

@@

A path is a sequence of edges. The edge-to-function mapping $f$ can be naturally extended to paths: if path $p$=$e_1 ; \cdots ; e_n$ then $f(p)=f(e_1) ; f(e_2) ; \cdots ; f(e_n)$.
%
We write $\partial(p)^{+}$ for $p$'s start node, $\partial(p)^{-}$ for its end node, and $\partial(p)$ to denote the pair
$(\partial(p)^{+}, \partial(p)^{-})$.

@@

parallelPairs $\gets$ \{\}\;
\For{src in existingGraph.Nodes}{
    \For{snk in existingGraph}{
        \try{}{
            \tcc{Use any standard path finding algorithm such as BFS to find a path in the existing graph from the specified source to sink.}
            Path pathWithNewEdge $\gets$ FindPath(
                sourceNode: src,
                sinkNode: newEdge.Source) +\;
            newEdge +\;
            FindPath(
                sourceNode: newEdge.Sink, 
                sinkNode: snk)\;
            \If{src == snk}{
                \tcc{Assign the nullary path from src to snk.}
                pathInOldGraph $\gets$ src\;
            }
            \Else(){
                Path pathInOldGraph $\gets$ FindPath(
                    sourceNode: src, 
                    sinkNode: snk)\;
            }
            parallelPairs.add((pathInOldGraph, pathWithNewEdge))\;
        }
        \catch{PathFindingFailedException}{
           \tcc{No comparable pairs from node src to node snk that need to be checked}
           continue\;
        }
    }
}
\KwRet{parallelPairs}\;
\caption{Online polynomial time algorithm to find parallel pair set.}
\label{algo_online_polynomial}
\end{algorithm}

@@

We conclude that verifying a comparable pair of paths with end points ($P_1$, $P_2$) implies the verification of all path pairs ($Q_1$, $Q_2$) such that $Q_1$ is a successor of $P_1$ and $P_2$ is a successor of $Q_2$. A successor $S$ to node $N$ is any node such that there exists a path from $N$ to $S$. Nodes are also their own successors and predecessors.
The rule effectively places an ordering on the informativeness of path pairs based on their terminal nodes.

@@

\begin{table}
    \caption{Computation time for 9-node graph of density 0.4, averaged over ten runs.}
\begin{tabular}{cc}
    \toprule
    Algorithm & Average seconds of computation \\
    \midrule
    Na\"{i}ve baseline & 0.77 \\
    Two Flip tolerant & 0.075 \\
    Batch algorithm & 7.55 \\
    ~\ref{algo_online_polynomial} & 0.0038 \\
    ~\ref{algo_online_minimal} & 0.00086 \\
    \bottomrule
\end{tabular}
\label{tab:times}
\end{table}

\begin{figure*}
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_OptimalSet.pdf}
      \caption{~\ref{algo_online_minimal}.}
      \label{fig:sfigOptimalTvsS}
    \end{subfigure}
    %
    \hspace{-20pt}
    %
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_Polynomial.pdf}
      \caption{~\ref{algo_online_polynomial}.}
      \label{fig:sfigPolynomialTvsS}
    \end{subfigure}
    %
    \hspace{-20pt}
    %
    \begin{subfigure}{0.35\linewidth}
      \includegraphics[width=\linewidth,trim=0 0 5 30,clip]{timeVsSize_10_NaiveChecker.pdf}
      \caption{Naive baseline.}
      \label{fig:sfigNaiveTvsS}
    \end{subfigure}

@@

\begin{figure}
    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{variance_10_OptimalSet.pdf}
      \caption{~\ref{algo_online_minimal}.}
      \label{fig:sfigOptimalSpread}
    \end{subfigure}

@@

We define density to be the ratio of the number of edges in the graph to the total possible number of edges (which is $|V|^2$, where $|V|$ is the number of nodes).
Run time relates to the density of edges in the input graph.
The degree of the effect differs with the algorithms, as Figure~\ref{fig:timeVsSize} shows.
Generally, denser graphs entail longer computation time.
For the batch algorithm we use lower densities since the input graph must be acyclic. This puts an upper bound on density that approaches 0.5 in large graphs.

@@

@@

Section ~\ref{sec:batchBaseline} describes Murota's solution to efficiently finding the minimal set of path pairs that need to be compared to check if a given acyclic graph commutes ~\cite{commutative}.
We did not find any other work that solves the question of verifying that diagrams commute.
\balance
However, the question of commuting does come up in programming languages with implicit type conversion.
Gator ~\cite{gator}, as described in Section ~\ref{sec:gator}, supports automatic type conversion between geometry types.
The language implements some restrictions to eliminate obvious cases of non-commuting graphs, but does not verify that defined graphs commute, allowing scope for non-commuting graph definitions.
Frink ~\cite{frink} is a language that supports automatic conversion between units and infinite precision floating point numbers.
It does not appear to support the implicit definition of conversion between units \xxx[AK]{Need to spend some time confirming this} but if extended to do so, would need to contend with the problem of commuting graphs.
The same is true for F\# which has support for units of measure ~\cite{unitsOfMeasure} and Ada's GNAT compiler~\cite{gnat}.

\section{Conclusion}
